---
title: 'STAT/CSSS 564: Assignment 4'
author: Jeff Arnold & Sheridan Grant
output: html_document
date: "May 14, 2017"
---

## Instructions

1. Fork this repository to your account
2. Edit the file `solutions.Rmd` with your solutions to the problems.
3. Submit a pull request to have it graded. Include an up-to-date knitted HTML or PDF file.

For updates and questions follow the Slack channel: [#assignment4](https://uwcsss564.slack.com/messages/C5DBV8266).

This assignment will require the following R packages:
```{r, message=FALSE}
library("rstan")
library("rstanarm")
```

And you might as well use all your cores:
```{r}
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## Problem 1: Prediction with Bayesian Penalized Regression
In this problem, you will perform a prediction test similar to May 11th's lab. You will compare four different shrinkage methods' prediction performance on a data set from the same [paper](https://www.cambridge.org/core/services/aop-cambridge-core/content/view/11866974EE2888D4A2988309FC6B602F/S1047198700013838a.pdf/improving_predictions_using_ensemble_bayesian_model_averaging.pdf). This data set contains a monthly record of whether or not a country was experiencing a violent insurgency from 1997 to 2010 for a number of countries, mostly in Asia and the South Pacific.

Your goal is to develop a model that predicts, as well as possible, future violent insurgencies in India from past ones. The higher-ups at the Department of Defense come to you with two specific objectives:
1. Choose a model that predicts the next month's insurgency status. Because this model is for such short-term use, a false negative (predicting no insurgency when there in fact will be one) should be considered 10 times as bad as a false positive.
2. Choose a model that predicts the next twelve months' insurgency status. For this model, either type of error is equally bad.

They are concerned about prompt availability of all data in the future, so they ask your model to include just 6 variables: the number of minority groups in the country ("nminorities"), GDP per capita ("gdppc"), the amount of anti-government violence ("antiGovtviolence"), the degree of cooperation with the USA ("USAcoop"), the degree of conflict with the USA ("USAconf"), and how near the next major election is ("nextelection").

```{r}
icews <- read.csv('data/icews.csv')
india <- icews[icews$country == "India",-1]

# The data
y <- india$insurgency
x <- india[c('nminorities', 'gdppc', 'antiGovtviolence', 'USAcoop', 'USAconf', 'nextelection')]
n <- length(y)
p <- dim(x)[2]

# What preprocessing of the data is necessary?
x$gdppc <- log(x$gdppc)
x$nextelection <- log(x$nextelection)
# TODO: preprocess x and y more
# y <- (y - mean(y))/sd(y)
for (i in 1:p) {
  x[,i] <- (x[,i] - mean(x[,i]))/sd(x[,i])
}

df <- cbind(y, x)
colnames(df)[1] <- 'insurgency'

# Example model fit
lmod <- glm(insurgency ~ ., family = binomial(), data = df)
lasso_mod <- stan_glm(insurgency ~ ., family = binomial(), data = df, prior = lasso())
```


*Note: it is justifiable to only base predictive performance on the last few years' data, since one might assume that conditions will change rapidly enough that old insurgencies will provide relatively little information. This also gives you an excuse to not spend forever waiting for code to run, which unfortunately you do not actually have in academia*


## Problem 2: Regularization Paths
In this problem, you will compare the regularization paths of models with various types of shrinkage priors, as well as frequentist LASSO and ridge regression models (similar to the beginning of the May 11 lab).

Can you use the type of prior distribution imposed on the regression coefficients to explain the shape of the regularization paths? [This paper](https://arxiv.org/pdf/1502.00560.pdf) may be of some help.